Hijacking Simulators with Universal Probabilistic Programming

### 2019-07-05 paper/learning/notes.tex
added more notes

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{graphicx}
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{listings}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{pgfplots}
\usepackage{tikz,pgf}
\usepackage{todonotes}
\usepackage{tikz-cd}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{wrapfig}

@@

One challenge in performing inference in such models is the combinatorial increase in traces, \emph{paths}
that can be taken within the given probabilistic model. This is because each member of the population has 
a bounded, but large number of decisions to make. In addition to this, each member of the population can interact
with other members of the population. To model a single population member particle-based methods such as 
sequential monte carlo (SMC) provide an ability to explore multiple paths

As updates are made sequentially MCMC methods typically are not-well suited this type of recusrive
estimation problem. As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
Likewise, importance sampling is not well suited either, as we cannot reuse the samples and weights for a time $t$ $\{x^{i}_{0:t}, \tilde{w}^{i}_{t}\}^{N}_{i=1}$ 
to sample from $(x_{0:t_1}| y_{0:t+1})$.

@@

\subsection{Prior based sampling}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{graphicx}
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{listings}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{pgfplots}
\usepackage{tikz,pgf}
\usepackage{todonotes}
\usepackage{tikz-cd}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{wrapfig}
\usepackage{amsmath}

@@

One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
that can be taken within the given probabilistic model.
This is because each member of the population has 
a bounded, but large number of decisions to make. 
In addition to this, each member of the population can interact
with other members of the population.
To model a single population member quantum-based methods leveraging 
linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
the whole program, \emph{state}, in one iteration.
 Unfortunately, analytically writing the functional 
form of the state is, in this instance, impossible.   
The usability of current classical inference methods in this context is also problematic. 
However, by combining hijacking and amortizing parts of the simulator, in particular rejection sampling 
loops, we can not only correctly perform inference in such models, as the rejection sampling loops are 
now replaced with learnt functions, but we can also perform inference much more efficiently. 
This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
construct the density and as they are amortized evaluation is cheap and only has to be performed once per call.
This both reduces memory consumption and running time.

As updates are made sequentially MCMC methods typically are not-well suited this type of recursive
estimation problem. 
As each time we get a new data point $y_{t+1}$ we have to run new MCMC simulations 
for $P(x_{0:t_1}| y_{0:t+1})$, we also cannot directly reuse the samples $\{x^{i}_{0:t}\}$. 
Likewise, importance sampling is not well suited either, as we cannot reuse the samples and weights for a time $t$ $\{x^{i}_{0:t}, \tilde{w}^{i}_{t}\}^{N}_{i=1}$ 
to sample from $(x_{0:t_1}| y_{0:t+1})$.
 However, as we can only run the simulator $\texttt{forward()}$ we cannot return back to a previous state and re-run. So adapting existing inference techniques 
 is the only viable option.

@@

\subsection{Prior based sampling}
In prior based sampling we take our existing program and look directly at the product of all the raw sample calls, denoted $f_{a_i}(x_i | \eta)$ and 
construct a proposal that is is dependent on those. In this sampling scheme our proposal is defined as:
\begin{equation}
  q(x_{1:n_{x}}) = 
  \begin{cases}
    \prod_{j=1}^{n_x} f_{a_i}(x_i | \eta) \hspace{0.1cm} \texttt{if} \hspace{0.1cm} \mathcal{B}(x_{1:n_{x})=1 \\
    0
  \end{cases}
\end{equation}
However, this makes use of no conditioning and is not sufficient for performing inference in
complex, population-based simulations.

\section{Non-prior Based Sampling}

### 2019-07-05 paper/learning/thesystem.tex
added more notes

 % STATE EPC
 \node[state,
  right of=PyProb,
  yshift = -0cm,
  node distance=5cm,
  anchor=center] (EPC) 
 {%
 \begin{tabular}{l}
  \textbf{Program density}\\
  \parbox{4cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$}
 \end{tabular}
 };

  % STATE OUTPUT
  \node[state,
  right of=System,
  yshift = 0cm,
  node distance=5cm,
  anchor=center] (OUTPUT) 
 {%
 \begin{tabular}{l}
  \textbf{Program Output}\\
  \parbox{4cm}{$\Omega = \{P(x_{1:n_{x}}), w\}$}
 \end{tabular}
 };

 % draw the paths and and print some Text below/above the graph
 \path (Sample) 	edge[bend left=20]  node[anchor=south,above]{$f_{a_{j}} ; \eta_{j}$}(PyProb)
 (OBSERVE)     	edge[bend right=20] node[anchor=south,above]{$g_{b_{i}},\phi_i , y_i $} (System)
 (PyProb)       	edge [bend left=20] node[anchor=south,above]{$g_{b_{i}}(y_i, \phi_i)$} (EPC)
 (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$ \emph{returned}} (OBSERVE)
 (System)       	edge [bend right=10]                                          (EPC)
 (EPC)              edge                                                    (OUTPUT)
 (System)  	edge   node[anchor=north,above]{\emph{Control}} (OBSERVE);
%  (System)  	edge[loop below]    node[anchor=north,below]{$SC_n\neq 0$} (System)
%  (System)  	edge                node[anchor=left,right]{$SC_n = 0$} (PyProb);

>>>>

 % STATE EPC
 \node[state,
  right of=System,
  yshift = 1cm,
  node distance=5cm,
  anchor=center] (EPC) 
 {%
 \begin{tabular}{l}
  \textbf{Program density}\\
  \parbox{5cm}{P(\mathcal{T}= $x_{1:n_x}| \lambda)$\\
  \{$x_{1:n_{x}}\}$,$w= \prod^{n_{y}}_{i=1}g_b_i (y_i | \psi_i)}$}
 \end{tabular}
 };

  % STATE OUTPUT
  \node[state,
  above of=EPC,
  yshift = -1cm,
  node distance=5cm,
  anchor=center] (OUTPUT) 
 {%
 \begin{tabular}{l}
  \textbf{Program Output}\\
  \parbox{5cm}{$\Omega = \{P_{r}(x_{1:n_{x}}), w\}$}
 \end{tabular}
 };

 % draw the paths and and print some Text below/above the graph
 \path (Sample) 	edge[bend left=20]  node[anchor=south,above]{$f_{a_{j}} ; \eta_{j}$}(PyProb)
 (OBSERVE)     	edge[bend right=20] node[anchor=south,above]{$g_{b_{i}},\phi_i , y_i $} (System)
 (System)       	edge [bend right=20] node[anchor=south,below]{$g_{b_{i}}(y_i, \phi_i)$} (EPC)
 (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$} (Sample)
 (PyProb)       edge [bend left=9] node[anchor=west, above]{$x_{i}$} (EPC)
%  (System)       	edge [bend left=10]                                          (EPC)
 (EPC)              edge                                                    (OUTPUT)
 (System)  	edge [bend right=10]  node[anchor=north,above]{\emph{Control}} (OBSERVE)
 (Sample)   edge [bend left= 20] (OUTPUT) ;
%  (System)  	edge[loop below]    node[anchor=north,below]{$SC_n\neq 0$} (System)
%  (System)  	edge                node[anchor=left,right]{$SC_n = 0$} (PyProb);
